---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "50%",
  fig.align = "center"
)
```

# Fast Sampling from Multivariate Normal distribution with Toeplitz-structured covariance matrix (MVToep)

<!-- badges: start -->

[![R-CMD-check](https://github.com/JaeHoanKim/MVToep/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/JaeHoanKim/MVToep/actions/workflows/R-CMD-check.yaml) [![Codecov test coverage](https://codecov.io/gh/JaeHoanKim/MVToep/branch/master/graph/badge.svg)](https://codecov.io/gh/JaeHoanKim/MVToep?branch=master)

<!-- badges: end -->

## Intended use 

Multivariate normal (Gaussian) sampling from the Toeplitz-structured covariance matrix is in diverse fields: from obtaining the approximate solution of the stochastic partial differential equation (see [1]) to directly testing the performance in the simulation studies (see [2]). Therefore, this package intends to efficiently draw samples from the Toeplitz-structured covariance matrix, with several user-friendly functions. For handy usage, it provides simple implementations for sampling from the regular grid points with radial basis function (RBF) and Matern kernels, with the covariance matrix of the autoregressive model. The main idea for this algorithm is suggested on [3].

## Installation

You can install the development version of MVToep package by: 

``` r
# install.packages("devtools")
devtools::install_github("JaeHoanKim/MVToep")
```

Then, use 

```
library(MVToep)
``` 

## Example

The following is a basic example which shows how to use ```rmvToep``` function.

To use the ```rmvToep``` function, a symmetric and Toeplitz-structured covariance matrix should be provided. ```mvToep``` package provides one class of covariance matrix that satisfies these conditions in ```Sigma.AR.order``` function, which is the covariance matrix of the autoregressive (AR) model. One can control the number of nonzero elements in this matrix by changing the ```order``` argument.


```{r example-rmvToep}
library(MVToep)
Sigma = Sigma.AR.order(0.7, 200, order = 5)
print(Sigma[1:8, 1:8])
```

Before applying ```rmvToep``` function on ```Sigma```, one can check the availability of this algorithm by using ```nnd.C.Toep``` function. This function returns the comment whether one can apply ```rmvToep``` function or not. ```Sigma``` should be positive semi-definite matrix to apply ```rmvToep``` function.

```{r nnd check - positive}
nnd.C.Toep(Sigma)
```

Here is the example in which one cannot apply ```rmvToep``` function:

```{r nnd check - negative}
Sigma2 = Sigma.AR.order(0.7, 200, order = 1)
nnd.C.Toep(Sigma2)
```

The example below illustrates the sampling result from ```rmvToep``` function.

```{r}
sample1 = rmvToep(500, Sigma)
plot(sample1[, 1], sample1[, 2])
(cor(sample1[, 1], sample1[, 2]))
```

Simply saying, ```order``` argument indicates the nonzero bandwidth of the matrix. Under the AR model, the correlation coefficient of the neighborhood is ```rho```, which matches the above result.

```{r}
plot(sample1[, 1], sample1[, 21])
(cor(sample1[, 1], sample1[, 21]))
```

In contrast, two indexes located far from each other show negligible correlation.

```rmvMat``` function works in a simpler way. This is the function to draw multivariate normal samples under Matern covariance kernel at a regular grid, which can be directly applied to obtain the approximate solution of the stochastic partial differential equation (SPDE). The object ```result_Mat``` contains two arguments as a separate list. The first list is the multivariate samples, and the second the used covariance matrix. One can use ```?MatK``` to check how the covariance matrix is calculated.

```{r example-rmvMat}
grid = c(0:300)/50
result_Mat = rmvMat(n = 40, grid, rho = 3, nu = 0.5, tau = 2)
sample2 = result_Mat[[1]]
```

By the visualization below, one can indirectly check that each row of the ```sample2``` indicates the Gaussian process with Matern covariance matrix.

```{r}
plot(sample2[1, ], ylim = c(-6, 6))
for (i in 1:7){
   lines(sample2[i, ], col = i)
}
```

```rmvRBF``` works similarly as ```rmvMat``` function. The only difference is the required arguments for the kernel setting. It is worth noting that rmvRBF is more likely to suffer from the ill-condition problem of the covariance matrix due to its nature. One can use ```?RBFK``` to check how the covariance matrix is calculated.

```{r}
grid = c(0:13)/13
result_RBF = rmvRBF(n = 40, grid, l = 0.1, tau = 1)
sample3 = result_RBF[[1]]
plot(sample3[1, ], ylim = c(-3, 3))
for (i in 1:7){
   lines(sample3[i, ], col = i)
}
```





## References

- Lindgren, Finn, Håvard Rue, and Johan Lindström. "An explicit link between Gaussian fields and Gaussian Markov random fields: the stochastic partial differential equation approach." Journal of the Royal Statistical Society: Series B (Statistical Methodology) 73.4 (2011): 423-498.

- Kim, Jaehoan, Hoyoung Park, and Junyong Park. "High dimensional discriminant rules with shrinkage estimators of covariance matrix and mean vector." arXiv preprint arXiv:2211.15063 (2022).

- Wood, Andrew TA, and Grace Chan. "Simulation of stationary Gaussian processes in [0, 1]^d." Journal of computational and graphical statistics 3.4 (1994): 409-432.



